---
title: "Data Wrangling with R"
subtitle: "IBM 3202 Section 02"
author: "Mia Polito"
date: "2026-03-11"
format:
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-location: right-body
        embed-resources: true
execute: 
  freeze: auto
---

> **Note**: This codebook is based on the content of the lecture notes published on <https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/>

# Overview

## Learning Outcomes

By the end of this module, you will be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions. (Next module)
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with mutate() and case_when().
-   Use the pipe operator to shape the data to prepare for analysis and visualization.

::: callout-tip
To use multi-cursor editing feature, do the following steps: 1. Select the text you want to edit. 2. Press 'shift + alt/option + i' to place a cursor at the end of each selected line. 3. You can now edit all the selected text simultaneously.
:::

# Introduction to Data Wrangling

## Loading Packages

**Note**:

-   To add a `code chunk`, you can use the keyboard shortcut `Ctrl + Alt + I` (Windows/Linux) or `Cmd + Option + I` (Mac) in RStudio.
-  In visual mode: `Ctrl +Alt +I` (Windows/Linux) or `Cmd + Option + I` (Mac) in RStudio.

-  In source mode: `Ctrl + Shift + I` (Windows/Linux) or `Cmd + Shift +I` (Mac) in RStudio.

```{r}
# install.packages("tidyverse")
# install.packages("nycflights13")
library(tidyverse)
library(nycflights13)
```

## Tibbles vs Data Frames

```{r}
class(mtcars)

cars.tib <- mtcars |> 
  rownames_to_column(var = "model") |> 
  as_tibble()

cars.tib
class(cars.tib)
typeof(cars.tib)
```

## Data Wrangling with dplyr

```{r}
data()
```

**Data sets in the package `nycflights13`:**

-  airlines         Airline names.
-  airports         Airport metadata
-  flights          Flights data
-  planes           Plane metadata.
-  weather          Hourly weather data

```{r}
flights_df <- flights

flights_df |> 
  count(year, month) |>
  arrange(desc(n)) |>
  mutate(month = factor(month)) |>
  mutate(month = fct_reorder(month, n)) |>
  ggplot(aes(month, n, fill = year)) +
  geom_col(show.legend = FALSE) +
   coord_flip()
  labs(x = "Month",
       y = "# of Flights",
       title = "# of flights by Month during year 2013 at New York") +
  theme_bw()
```
